\documentclass[11pt]{article}

\input{preamble}

\begin{document}
{\bf \noindent Blue MOP 2009; Josh \\ Squares!}

\be

\ii For which primes $p$ is $-15$ a square?
\ii Prove that the equation $x^2+5xy-3y^2 = 131$ has no integer solutions.
\ii Suppose $p=12^n+1$ is prime.  Under what circumstances does $p|5^{\frac{p-1}2}+1$?

% \ii Given $D$ squarefree and a solution $(x_1,y_1)$ to $x^2-Dy^2=4$ in positive integers with minimal $x$-value, show that all other solutions may be written as \[\frac{x_n+y_n\sqrt{D}}2 = \Lp\frac{x_1+y_1\sqrt D}2\Rp^n.\]
% \ii Given $D$ squarefree and a solution $(x_1,y_1)$ to $x^2-Dy^2=-1$ in positive integers with minimal $x$-value, find all other solutions in positive integers.
\ii A pentagonal number is a number of the form $\frac{n(3n-1)}2$, where $n$ is a positive integer.  Find the first pentagonal number greater than 1 that is also a perfect square.
\ii Find all pairs of consecutive cubes whose difference is a perfect square.
\ii Find the smallest positive integer $n$ such that $19n+1$ and $95n+1$ are both integer squares.
\ii Consider the sequence of reals $(a_n)_{n=0}^\infty$ defined recursively by $a_1=1$ and \[a_{n+1} = 2a_n + \sqrt{3a_n^2-3}.\]  Prove that $a_n$ is an integer for all $n$.
\ii Prove that the equation \[x^3+y^3+z^3+t^3 = 1999\] has infinitely many integer solutions. % Start with $(10,10,-1,0)$ and deform to $(10-b,10+b,d-1,-d)$ or somesuch
\ii Determine all positive integers $n$ for which there exists an
integer $m$ such that $2^n-1$ is a divisor of $m^2 + 9$.
\ii (MOP '98) Find all pairs of positive integers $(x,y)$ satisfying the equation
\[
x^{2} + (x+1)^{2} + \cdots + (x+99)^{2} = y^{2} + (y+1)^{2} + \cdots
+ (y+100)^{2}.
\]
\ii (MOP '09) Find all solutions in nonnegative integers to \[2^m+7^n = 9^k.\]
\ii Find all solutions in positive integers to $3^x-2=y^2$.
\ii (MOP '97) Does there exist a row of Pascal's Triangle containing four distinct
elements $a, b, c$, and $d$ such that $b=2a$ and $d= 2c$?
\ii Prove that the equation \[x^3+y^3 = z^4 - t^2\] has infinitely many integer solutions with $\gcd(x,y,z,t) = 1$.
\ii (MOP '95) The integer $n$ is a square mod $p$ for every prime $p$. Prove that $n$ is a perfect square.

\ii Find all pairs of nonnegative integers $(m, n)$ such that
\[
(m+n-5)^2 = 9mn.
\]

\ee
\end{document}